---
title: "Introduction to Bulk RNAseq data analysis"
subtitle: Initial exploration of RNA-seq data
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: true
bibliography: ref.bib
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 5)
```

# 1. Introduction

In this section we will begin the process of analyzing the RNAseq data in R. In
the next section we will use
[DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html) for
differential analysis. A detailed analysis workflow, recommended by the authors
of DESeq2 can be found on [the Bionconductor
website](http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html).

Before embarking on the main analysis of the data, it is essential to do some
exploration of the raw data. We want to assess the patterns and 
characteristics of the data and compare these to what we expect from mRNAseq 
data and assess the data based on our knowledge of the experimental design. The
primary means of data explorations are summary statistics and visualisations.
In this session we will primarily concentrate on assessing if the patterns
in the raw data conform to what we know about the experimental design. This is
essential to identify problems such as batch effects, outlier samples and 
sample swaps.

Due to time constraints we are not able to cover all the ways we might do this,
so additional information on initial data exploration are available in the
[supplementary materials](S5_Additional_Data_Exploration.html).

In this session we will:

* import our counts into R  
* filter out unwanted genes  
* transform the data to mitigate the effects of variance  
* do some initial exploration of the raw count data using principle component 
analysis and hierarchical clustering  

# 2. Data import

First, let's load all the packages we will need to analyse the data.

```{r packages, message = FALSE}
library(tximport)
library(DESeq2)
library(tidyverse)
```

## A brief description of the data set

The data for this tutorial comes from the paper [Transcriptomic Profiling of 
Mouse Brain During Acute and Chronic Infections by *Toxoplasma gondii*
Oocysts](https://www.frontiersin.org/articles/10.3389/fmicb.2020.570903/full) 
[@Hu2020]. The raw data (sequence reads) can be downloaded from the [NCBI Short 
Read Archive](https://www.ncbi.nlm.nih.gov/sra) under project number 
**PRJNA483261**.

Please see extended material for instructions on [downloading raw files from
SRA](S1_Getting_raw_reads_from_SRA.html).

This study examines changes in the gene expression profile in mouse brain in 
response to infection with the protozoan *Toxoplasma gondii*. The authors
performed transcriptome analysis on samples from infected and uninfected mice at 
two time points, 11 days post infection and 33 days post infection. For each 
sample group there are 3 biological replicates. This effectively makes this a 
two factor study with two groups in each factor:

* Status: Infected/Uninfected
* Time Point: 11 dpi/33 dpi

## Reading in the sample metadata

The `SampleInfo.txt` file contains basic information about the samples that we will
need for the analysis today: name, cell type, status.

```{r loadSampleInfo, message = FALSE}
# Read the sample information into a data frame
sampleinfo <- read_tsv("data/samplesheet.tsv", col_types = c("cccc"))
arrange(sampleinfo, Status, TimePoint, Replicate)
```

## Reading in the count data

Salmon [@Patro2017] was used to quantify gene expression from raw reads against
the Ensembl transcriptome GRCm38 version 102 (as described in 
[the previous session](03_Quantification_with_Salmon_practical.html)).

First we need to read the data into R from the `quant.sf` files under the 
*salmon* directory. To do this we use the `tximport` function. We need to create
a named vector in which the values are the paths to the `quant.sf` files and the
names are sample names that we want in the column headers - these should match
the sample names in our `sampleinfo` table.

The Salmon quantification results are per transcript, we'll want to summarise
to gene level. To this we need a table that relates transcript IDs to gene IDs.

```{r readSalmon}
files <- file.path("salmon", sampleinfo$SampleName, "quant.sf")
files <- set_names(files, sampleinfo$SampleName)
tx2gene <- read_tsv("references/tx2gene.tsv")

txi <- tximport(files, type = "salmon", tx2gene = tx2gene)
str(txi)
head(txi$counts)
```

Save the `txi` object for use in later sessions.

```{r saveData, eval = FALSE}
saveRDS(txi, file = "salmon_outputs/txi.rds")
```

### A quick intro to `dplyr`

One of the most complex aspects of learning to work with data in `R` is 
getting to grips with subsetting and manipulating data tables. The package 
`dplyr` [@Wickham2018] was developed to make this process more intuitive than it
is using standard base `R` processes. 

In particular we will use the commands:

* `select` to select columns from a table
* `filter` to filter rows based on the contents of a column in the table
* `rename` to rename columns

We will encounter a few more `dplyr` commands during the course, we will explain
their use as we come to them.

If you are familiar with R but not `dplyr` or `tidyverse` then we have a very
brief introduction [here](S4_A_brief_intro_to_dplyr.html). A more detailed 
introduction can be found in our [online R course](https://bioinformatics-core-shared-training.github.io/r-intro/week4.html)

# 3. Prepare count matrix

## Create a raw counts matrix for data exploration

DESeq2 will use the txi object directly but we will need a counts matrix to
do the data exploration.

```{r rawCounts}
rawCounts <- round(txi$counts, 0) 
```

## Filtering the genes

Many, if not most, of the genes in our annotation will not have been detected at
meaningful levels in our samples - very low counts are most likely technical
noise rather than biology. For the purposes of visualization it is important to
remove the genes that are not expressed in order to avoid them dominating the
patterns that we observe.

The level at which you filter at this stage will not effect the differential
expression analysis. The cutoff used for filtering is a balance between removing
noise and keeping biologically relevant information. A common approach is to
remove genes that have less than a certain number of reads across all samples.
The exact level is arbitrary and will depend to some extent on nature of the
dataset (overall read depth per sample, number of samples, balance of read depth
between samples etc). We will keep all genes where the total number of reads
across all samples is greater than 5.

```{r filterGenes}
# check dimension of count matrix
dim(rawCounts)
# for each gene, compute total count and compare to threshold
# keeping outcome in vector of 'logicals' (ie TRUE or FALSE, or NA)
keep <- rowSums(rawCounts) > 5
# summary of test outcome: number of genes in each class:
table(keep, useNA = "always") 
# subset genes where test was TRUE
filtCounts <- rawCounts[keep,]
# check dimension of new count matrix
dim(filtCounts)
```

# 4. Count distribution and Data transformations

Differential expression calculations with DESeq2 uses raw read counts as input,
but for visualization purposes we use transformed counts.

## Raw counts 

Why not raw counts? Two issues:

* The range of values in raw counts is very large with many small values and a few
  genes with very large values. This can make it difficult to see patterns in the
  data.

```{r raw_summary}
summary(filtCounts)
```

```{r raw_boxplot}
# few outliers affect distribution visualization
boxplot(filtCounts, main = 'Raw counts', las = 2)
```

* Variance increases with mean gene expression, this has impact on assessing
  the relationships, e.g. by clustering.

```{r raw_mean_vs_sd}
# Raw counts mean expression Vs standard Deviation (SD)
plot(rowMeans(filtCounts), rowSds(filtCounts), 
     main = 'Raw counts: sd vs mean', 
     xlim = c(0, 10000),
     ylim = c(0, 5000))
```

## Data transformation

To avoid problems posed by raw counts, they can be
[transformed](http://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#data-transformations-and-visualization).
A simple log2 transformation can be used to overcome the issue of the range of
values. Note, when using a log transformation, it is important to add a small
"pseudocount" to the data to avoid taking the log of zero.

```{r log2}
logCounts <- log2(filtCounts + 1)
boxplot(logCounts, main = 'Log2 counts', las = 2)
```

However, this transformation does not account for the variance-mean
relationship. DESeq2 provides two additional functions for transforming the
data:

* `VST` : variance stabilizing transformation
* `rlog` : regularized log transformation

As well as log2 transforming the data, both transformations produce data which
has been normalized with respect to library size and deal with the mean-variance
relationship. The effects of the two transformations are similar. `rlog` is
preferred when there is a large difference in library size between samples,
however, it is considerably slower than `VST` and is not recommended for large
datasets. For more information on the differences between the two
transformations see the
[paper](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8)
and the DESeq2 vignette.

Our data set is small, so we will use `rlog` for the transformation.

```{r rlog}
rlogcounts <- rlog(filtCounts)
boxplot(rlogcounts, main = 'rlog counts', las = 2)
```

# 5. Principal Component Analysis

A principal component analysis (PCA) is an example of an unsupervised analysis,
where we don't specify the grouping of the samples. If the experiment is well
controlled and has worked well, we should find that replicate samples cluster 
closely, whilst the greatest sources of variation in the data should be between
treatments/sample groups. It is also an incredibly useful tool for checking for 
outliers and batch effects.

To run the PCA we should first normalise our data for library size and transform
to a log scale. DESeq2 provides two separate commands to do this (`vst` and
`rlog`). Here we will use the command `rlog`. `rlog` performs a log2 scale
transformation in a way that compensates for differences between samples for
genes with low read count and also normalizes between samples for library size.

You can read more about `rlog`, its alternative `vst` and the comparison
between the two 
[here](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-data-transformations).

To plot the PCA results we will use the `autoplot` function from the
`ggfortify` package [@Tang2016]. `ggfortify` is built on top of `ggplot2` and
is able to recognise common statistical objects such as PCA results or linear
model results and automatically generate summary plot of the results in an
appropriate manner.

```{r pcaPlot, message = FALSE, fig.width = 6.5, fig.height = 5, fig.align = "center"}
library(ggfortify)

rlogcounts <- rlog(filtCounts)

# run PCA
pcDat <- prcomp(t(rlogcounts))
# plot PCA
autoplot(pcDat)
```

We can use colour and shape to identify the Cell Type and the Status of each
sample.

```{r pcaPlotWiColor, message = FALSE, fig.width = 6.5, fig.height = 5, fig.align = "center"}
autoplot(pcDat,
         data = sampleinfo, 
         colour = "Status", 
         shape = "TimePoint",
         size = 5)
```

## Exercise
>
> The plot we have generated shows us the first two principle components. This
> shows us the relationship between the samples according to the two greatest
> sources of variation. Sometime, particularly with more complex experiments 
> with more than two experimental factors, or where there might be confounding
> factors, it is helpful to look at more principle components.
>
> 1. Redraw the plot, but this time plot the 2nd principle component on the 
> x-axis and the 3rd prinicple component on the y axis. To find out how to do
> the consult the help page for the `prcomp` data method for the `autoplot`
> function: `?autoplot.prcomp`.

```{r solutionChallenge1}

```

> ### Discussion: What do the PCA plots tell us about our samples?

<!-- amount and source of variation, sample swap, interaction -->

Let's identify these samples. The package `ggrepel` allows us to add text to 
the plot, but ensures that points that are close together don't have their
labels overlapping (they *repel* each other).

```{r badSamples, fig.width = 6.5, fig.height = 5, fig.align = "center"}
library(ggrepel)

# setting shape to FALSE causes the plot to default to using the labels instead of points
autoplot(pcDat,
         data = sampleinfo,  
         colour = "Status", 
         shape = "TimePoint",
         size = 5) +
    geom_text_repel(aes(x = PC1, y = PC2, label = SampleName), box.padding = 0.8)
```

The mislabelled samples are *SRR7657882*, which is labelled as *Infected* but
should be *Uninfected*, and *SRR7657873*, which is labelled as *Uninfected* but
should be *Infected*. Let's fix the sample sheet.

We're going to use another `dplyr` command `mutate`. 

```{r correctSampleSheet}
sampleinfo <- mutate(sampleinfo, Status = case_when(
                                          SampleName=="SRR7657882" ~ "Uninfected",
                                          SampleName=="SRR7657873" ~ "Infected", 
                                          TRUE ~ Status))
```

...and export it so that we have the correct version for later use.

```{r, exportSampleSheet, eval = FALSE}
write_tsv(sampleinfo, "results/SampleInfo_Corrected.txt")
```

Let's look at the PCA now.

```{r correctedPCA, fig.width = 6.5, fig.height = 5, fig.align = "center"}
autoplot(pcDat,
         data = sampleinfo, 
         colour = "Status", 
         shape = "TimePoint",
         size = 5)
```

Replicate samples from the same group cluster together in the plot, while 
samples from different groups form separate clusters. This indicates that the
differences between groups are larger than those within groups.
The biological signal of interest is stronger than the noise (biological and 
technical) and can be detected. 

Also, there appears to be a strong difference between days 11 and 33 post 
infection for the infected group, but the day 11 and day 33 samples for the
uninfected are mixed together.

Clustering in the PCA plot can be used to motivate changes to the design
matrix in light of potential batch effects. For example, imagine that the
first replicate of each group was prepared at a separate time from the second
replicate. If the PCA plot showed separation of samples by time, it might be
worthwhile including time in the downstream analysis to account for the
time-based effect.

# 6. Hierachical clustering

Earlier, we used principle component analysis to assess sources of variation
in the data set and the relationship between the samples. Another method for
looking at the relationship between the samples can be to run hierarchical 
clustering based on the Euclidean distance between the samples. Hierarchical
clustering can often provide a clearer view of the clustering of the different
sample groups than other methods such as PCA.

We will use the package `ggdendro` to plot the clustering results using the
function `ggdendrogram`.

```{r}
library(ggdendro)
hclDat <-  t(rlogcounts) %>%
   dist(method = "euclidean") %>%
   hclust()
ggdendrogram(hclDat, rotate = TRUE)
```

We really need to add some information about the sample groups. The simplest way
to do this would be to replace the `labels` in the `hclust` object. Conveniently
the labels are stored in the `hclust` object in the same order as the columns
in our counts matrix, and therefore the same as the order of the rows in our
sample meta data table. We can just substitute in columns from the metadata.

```{r}
hclDat2 <- hclDat
hclDat2$labels <- str_c(sampleinfo$Status, ":", sampleinfo$TimePoint)
ggdendrogram(hclDat2, rotate = TRUE)
```

We can see from this that the infected and uninfected samples cluster separately
and that day 11 and day 33 samples cluster separately for infected samples, but
not for uninfected samples.

--------------------