---
title: "Introduction to Bulk RNAseq data analysis"
subtitle: Initial exploration of RNA-seq data - solutions
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, message = FALSE}
library(tximport)
library(DESeq2)
library(tidyverse)
library(ggfortify)
```

```{r prepareData, message=FALSE, include=FALSE}
# This chunk is just for the knitting
# Read the sample information into R
sampleinfo <- read_tsv("data/samplesheet.tsv", col_types = c("cccc"))
# Read the data into R
files <- file.path("salmon", sampleinfo$SampleName, "quant.sf")
files <- set_names(files, sampleinfo$SampleName)

tx2gene <- read_tsv("references/tx2gene.tsv")

txi <- tximport(files, type = "salmon", tx2gene = tx2gene)

rawCounts <- round(txi$counts, 0)
# filter
keep <- rowSums(rawCounts) > 5
filtCounts <- rawCounts[keep,]
# VST
vst_counts <- vst(filtCounts)
statusCols <- str_replace_all(sampleinfo$Status, c(Infected="red", Uninfected="orange"))
rlogcounts <- rlog(filtCounts)

# run PCA
pcDat <- prcomp(t(rlogcounts))
```

# 5. Principal Component Analysis

## Exercise
>
> The plot we have generated shows us the first two principle components. This
> shows us the relationship between the samples according to the two greatest
> sources of variation. Sometime, particularly with more complex experiments 
> with more than two experimental factors, or where there might be confounding
> factors, it is helpful to look at more principle components.
>
> 1. Redraw the plot, but this time plot the 2nd principle component on the 
> x-axis and the 3rd prinicple component on the y axis. To find out how to do
> the consult the help page for the `prcomp` data method for the `autoplot`
> function: `?autoplot.prcomp`.

```{r solutionExercise3}
autoplot(pcDat,
         data = sampleinfo, 
         colour="Status", 
         shape="TimePoint",
         x=2,
         y=3,
         size=5)
```