---
title: "Introduction to Bulk RNAseq data analysis"
subtitle: Initial exploration of RNA-seq data - solutions
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
layout: page
---

### Data 

```{r setup, message = FALSE}
library(tximport)
library(DESeq2)
library(tidyverse)
library(ggfortify)
```

```{r prepareData, message=FALSE, include=FALSE}
# This chunk is just for the knitting
# Read the sample information into R
sampleinfo <- read_tsv("data/samplesheet.tsv", col_types = c("cccc"))
# Read the data into R
files <- str_c("salmon/", sampleinfo$SampleName, "/quant.sf")
files <- set_names(files, sampleinfo$SampleName)

tx2gene <- read_tsv("references/tx2gene.tsv")

txi <- tximport(files, type = "salmon", tx2gene = tx2gene)

rawCounts <- round(txi$counts, 0)
# filter
keep <- rowSums(rawCounts) > 5
filtCounts <- rawCounts[keep,]
# VST
vst_counts <- vst(filtCounts)
statusCols <- str_replace_all(sampleinfo$Status, c(Infected="red", Uninfected="orange"))
rlogcounts <- rlog(filtCounts)

# run PCA
pcDat <- prcomp(t(rlogcounts))
```

### Exercise 1
>
> We have loaded in the raw counts here. These are what we need for the 
> differential expression analysis. For other investigations we might want 
> counts normalised to library size. `tximport` allows us to import 
> "transcript per million" (TPM) scaled counts instead.
>
> 1. Create a new object called `tpm` that contains length scaled TPM 
>    counts. You will need to add an extra argument to the command. Use the help
>    page to determine how you need to change the code: `?tximport`.

```{r, solutionExercise1}
tpm <- tximport(files, type = "salmon", tx2gene = tx2gene, 
                countsFromAbundance = "lengthScaledTPM")
```

### Exercise 2
>
> 1. Use the `DESeq2` function `rlog` to transform the count data. This function
> also normalises for library size.
> 2. Plot the count distribution boxplots with this data.
> How has this effected the count distributions?

```{r, message=FALSE}
rlogcounts <- rlog(filtCounts)

# Check distributions of samples using boxplots
boxplot(rlogcounts, 
        xlab="", 
        ylab="Log2(Counts)",
        las=2,
        col=statusCols)
# Let's add a blue horizontal line that corresponds to the median logCPM
abline(h=median(as.matrix(rlogcounts)), col="blue")
```

### Exercise 3
>
> The plot we have generated shows us the first two principle components. This
> shows us the relationship between the samples according to the two greatest
> sources of variation. Sometime, particularly with more complex experiments 
> with more than two experimental factors, or where there might be confounding
> factors, it is helpful to look at more principle components.
>
> 1. Redraw the plot, but this time plot the 2nd principle component on the 
> x-axis and the 3rd prinicple component on the y axis. To find out how to do
> the consult the help page for the `prcomp` data method for the `autoplot`
> function: `?autoplot.prcomp`.

```{r solutionExercise3}
autoplot(pcDat,
         data = sampleinfo, 
         colour="Status", 
         shape="TimePoint",
         x=2,
         y=3,
         size=5)
```