---
title: "Introduction to Bulk RNAseq data analysis"
subtitle: Annotation of Differential Expression Results
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
bibliography: ref.bib
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 4, fig.height = 3)
```

# Adding annotation to the DESeq2 results

We have a list of significantly deferentially expressed genes, but the only
annotation we can see is the Ensembl Gene ID, which is not very informative. 

There are a number of ways to add annotation. One method is to do this using a 
Bioconductor annotation package. These packages which are re-built every 
periodically with the latest annotations. These packages are listed on the 
[annotation section](http://bioconductor.org/packages/release/BiocViews.html#___AnnotationData) 
of the Bioconductor, and are installed in the same way as regular Bioconductor 
packages. 

Another approach is to use `biomaRt`, an interface to the 
[BioMart](http://www.biomart.org/) resource. Using BioMart ensures that you are 
able to get the latest annotations for the GeneIDs, and can match the version of
the gene annotation that was used for read counting.

A third method is to use `AnnotationHub`, this is like the bioconductor
packages but in an online database like `bioMaRt`. They keep them slightly more
up to date than the standard bioconductor packages and each time you use them
the results are cached on your machine. 

We could also simply download an annotation table from a suitable repository,
for example Gencode provided tables for human and mouse that map Ensembl IDs to
Gene Symbol and Entrez IDs. This may be the most straightforward method in some
cases, but it is not as flexible as the other methods.

Today we will demonstrate how to use `AnnotationHub`. A workflow for annotation
with biomaRt is included in the extended materials section accessible on the
course website.

First we will load the differential gene expression results obtained from the 
interaction model for Infected vs Uninfected at day 11.

```{r packages, message=FALSE, warning=FALSE}
library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)
library(DESeq2)
library(tidyverse)
```


```{r loadData}
results.d11 <- readRDS("RObjects/DESeqResults.interaction_d11.rds")
```


# Query the database

We need to get the correct database from `AnnotationHub`. We make the
instance of the database in our R session (the first time we do this it will
create a local cache on your machine so that repeat queries are very quick). 

```{r, makeOrgDb, message=F}
ah <- AnnotationHub()
ah
```

As you can see `ah` contains huge amounts of information and it is constantly
changing. This is why it gives us the snapshot date so we know when our cached
version is from. The `ah` object actually online contains pointers to where all
the information is online and we don't want to download all of them as it would
take a very long time and we don't need all of it.

The object is a vector and you can get information about a single resource by
indexing it with a single bracket `[` or actually download a resource with a
double bracket `[[`.

```{r, inspectOrgDb}
ah[1]
```

Now we can query the database for the annotation we need. We can use the
`query` function to search for the database we need by providing it with
a vector of search terms. In our case we will look for:

* The Ensembl database - from which we obtained our reference:`EnsDb` 
* The species - `Mus musculus`
* The specific Ensembl release version - `102`

Genomic and transcriptomic databases undergo a lot of changes and updates, so it
is important to note which version of the database you are using when you download
your reference. The current course materials use Ensembl release 102 (a little old,
but we will update soon).

```{r, queryDB}
dbResult <- query(ah, c("EnsDb", "Mus musculus", "102"))
dbResult[1]
```

This returns one record of a database in Annotation Hub that matches our
requirements. We can download it using the `[[` operator.

## Download the relevant resource

```{r, downloadDB, message=FALSE}
MouseEnsDb <- ah[[dbResult$ah_id]]
```

This database contains the entire gene annotation from Ensembl release 102 for
mouse. It includes descriptions of genes, transcripts, exons, UTRs etc. 

We can turn the whole gene-level annotation table into a data frame so we can
work with it using the tidyverse suite of tools.

```{r queryOrgDb, message=F}
annotations <- genes(MouseEnsDb, return.type = "data.frame")

t(annotations[1, ])
```

## Subset the annotation to the genes and columns of interest

We are just going to keep the gene symbol, the description, and also the Entrez ID (which we
will need later on for gene set testing).

```{r inspectAnnot}
annot <- annotations %>%
    select(gene_id, gene_name, description, entrezid) %>%
    filter(gene_id %in% rownames(results.d11))
```

**NOTE**: You may get an error with this command that looks like:

```{css, echo=FALSE}
.error {
   color: red;
}
```

```{class.source="error"}
Error in (function (classes, fdef, mtable)  : 
  unable to find an inherited method for function 'select' for signature '"data.frame"'
```

This is due to the `select` function from **dplyr** (part of **tidyverse**)
being masked by the `select` function from one of the annotation packages. This
will have happened because the annotation package was loaded after the
**tidyverse**. You can either restart your R session and reload the required
packages, this time being sure to load **tidyverse** last, or just use
`dplyr::select` to explicitly use the `select` function from **dplyr**:

```{r maskingSolutino, eval = FALSE}
annot <- annotations %>%
    dplyr::select(gene_id, gene_name, description, entrezid) %>%
    dplyr::filter(gene_id %in% rownames(results.d11))
```

## Missing annotations

Let's inspect the annotation.

```{r inspectAnnotMis}
length(annot$entrezid)
length(unique(annot$entrezid))
sum(is.na(annot$entrezid))
```

There nearly 3000 genes with missing Entrez IDs. Gene/transcript/protein IDs
mapping between different databases not always perfect. 

Although majority of IDs map between databases, small subset may not have
matching ID or may have more than one match. 
This is because feature identification algorithms, naming methodologies and
versions may differ among databases. For instance NCBI and HGNC give same ID
for different gene versions, whereas Ensembl assigned separate IDs for gene
versions. Read interesting discussion on
[biostars](https://www.biostars.org/p/119540/).

There are some Ensembl IDs with no EntrezID. These gene ids has no 
corresponding Entrez ID in the `EnsDb` database package. The 
Ensembl and Entrez databases don't match on a 1:1 level although they have 
started taking steps towards consolidating
[in recent years](https://m.ensembl.org/info/genome/genebuild/mane.html).

## Duplicated annotations

In addition to the missing Entrez, there are number of cases in which multiple
Ensembl IDs map to the same Entrez ID.

```{r inspectAnnotdup}
dupEntrez <- annot %>%
  filter(!is.na(entrezid)) %>%
  add_count(entrezid) %>%
  arrange(entrezid) %>%
  filter(n > 1)
head(dupEntrez)
```

In this case many of these genes also have the same gene name and description,
in these cases they lie on patches or haplotypes, for more information see this
video:

https://www.youtube.com/watch?v=sPE9j_Hw9HU

These duplicates could cause problems with downstream analysis and should be
resolved. Often there are not many of these and they can be manually checked and
resolved by looking at the gene annotations online, based on which you can make
a decision about keeping the annotation or removing one or all of the
duplicates. Another option for resolving duplicates is to keep the one with the
highest count in the DESeq2 results. There is no ideal solution here and it is
up to the researcher to decide what to do.

To save time we have already created an annotation in which the problems with
duplicate annotations have been resolved. We will load this annotation and use
it for the rest of the course.

```{r loadAnnot}
annot <- readRDS("RObjects/Ensembl_annotations.rds")
```

# Annotation our DE results

```{r addAnnotation, message=FALSE}
results.d11 <- as.data.frame(results.d11) %>%
    rownames_to_column("GeneID") %>%
    left_join(annot, "GeneID")
```

**NOTE**: If you had the issue earlier with `select` being masked and you used 
the `dplyr::select` solution, you will also have an issue `rename` and will need
to use `dplyr::rename`.

Finally we can output the annotation DE results using `write_tsv`.

```{r outputDEtables, eval=FALSE}
write_tsv(results.d11, "results.d11_Results_Annotated.txt")
```
