---
title: "Introduction to Bulk RNAseq data analysis"
subtitle: Differential Expression of RNA-seq data
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, echo=FALSE, message=FALSE}
options(tibble.print_max = 4, tibble.print_min = 4, max.print=40, 
        tibble.max_extra_cols=2)
library(tidyverse)
library(DESeq2)
```

### Exercise 1

> So far we have fitted a simple model considering just "Status", but in reality 
> we want to model the effects of both "Status" and "Time Point".
> 
> Let's start with the model with only main effects - an additive model with no 
> interaction. The main assumption here is that the effects of Status and the
> effects of Time Point are indepedent.
> 
> Recapitulate the above steps to generate a new DESeq2 object with
> the additive model. Then we will extract the results table as above.
> 
> #### Load the raw data, remembering to set the factor on the Status so that
> "Uninfected" will be set as the intercept:    
>   
```{r}
txi <- readRDS("RObjects/txi.rds")
sampleinfo <- read_tsv("data/samplesheet_corrected.tsv", col_types="cccc") %>% 
                   mutate(Status = fct_relevel(Status, "Uninfected"))
```
> 
> #### Create the model:
> 
```{r}
additive.model <- as.formula(~ TimePoint + Status)
```
> 
> #### Then build the DESeq from the raw data, the sample meta data and the model:
> 
```{r}
ddsObj.raw <- DESeqDataSetFromTximport(txi = txi,
                                      colData = sampleinfo,
                                      design = additive.model)
```
> 
> #### Filter the data set:
>
```{r}
keep <- rowSums(counts(ddsObj.raw)) > 5
ddsObj.filt <- ddsObj.raw[keep,]
```
>
> You are now ready to run the differential gene expression analysis
> Run the DESeq2 analysis
> 
> 1. Run the size factor estimation, dispersion estimation and modelling steps 
> using the `DESeq` command as above. 
>
```{r}
ddsObj <- DESeq(ddsObj.filt)
```

> 
> 2. Extract the default contrast using the `results` command into a new object 
> called `results.additive`
>
```{r}
results.additive <- results(ddsObj, alpha=0.05)
results.additive
```
>
> a) What contrast are these results for? If you have constructed the model 
>    correctly, then it should be the same as previous `results.simple`
Again this results table is for the contrast Infected v Uninfected.
> b) How many genes have an adjusted p-value of less than 0.05
```{r}
sum(results.additive$padj < 0.05, na.rm = TRUE)
```


### Exercise 2

> If we want a different contrast we can just pass the `results` function the 
> **name** of the contrast, as given by `resultsNames(ddsObj)`.
> Look at the help page for the `results` command to see how to do this.
>
> 1. Retrieve the results for the contrast of d33 versus d11. 
```{r}
results.d33vd11 <- results(ddsObj, name= "TimePoint_d33_vs_d11", alpha=0.05)
```

> 2. How many differentially expressed genes are there at FDR < 0.05?
```{r}
sum(results.d33vd11$padj < 0.05, na.rm = TRUE)
```

### Exercise 3
>
> When we looked at the PCA it did seem that an interaction model might be
> warranted. Let's test that.  
>  
> 1. Create a new DESeq2 object using a model with an interaction between 
>    TimePoint and Status. The model formula should be 
>
>    `~TimePoint + Status + TimePoint:Status`
>
>    where `TimePoint:Status` is the parameter for the interaction beteween 
>    TimePoint and Status.
>
> Note that `*` can be used as shortcut to add the interaction term,
> e.g. `~TimePoint * Status`, however, writing out in long form is clearer here.
>
> Remember to filter to remove uninformative genes.
>
```{r}
interaction.model <- as.formula(~ TimePoint * Status)
ddsObj.raw <- DESeqDataSetFromTximport(txi = txi,
                                      colData = sampleinfo,
                                      design = interaction.model)
keep <- rowSums(counts(ddsObj.raw)) > 5
ddsObj.filt <- ddsObj.raw[keep,]
```
> 2. Run the statistical analysis using the `DESeq` command and create a new
>    analysis object called `ddsObj.interaction`.
```{r}
ddsObj.interaction <- DESeq(ddsObj.filt)
```
> 3. Use the LRT to compare this to the simpler additive model 
>    (`~TimePoint + Status`)
```{r}
ddsObj.LRT <- DESeq(ddsObj.interaction, test="LRT", reduced=additive.model)
results.Interaction_v_Additive <- results(ddsObj.LRT)
```
> 4. Extract a table of results using `results`. 
> For how many genes is interaction model a better fit?  
```{r}
table(results.Interaction_v_Additive$padj < 0.05)
```

### Exercise 4
>
> Let's investigate the uninfected mice
>  
> 1. Extract the results for d33 v d11 for Uninfected mice. 
The the intercept is Uninfected mice at 11 days post infection, so the main 
effect `TimePoint_d33_vs_d11` is the result that we want.
```{r}
results.d33_v_d11_uninfected <- results(ddsObj.interaction, 
                                        name="TimePoint_d33_vs_d11",
                                        alpha = 0.05)
```
> How many genes have an adjusted p-value less than 0.05?  
```{r}
table(results.d33_v_d11_uninfected$padj < 0.05)
```
> Is this remarkable?  
Maybe not. Do we really expect vast gene expression differences between the 
brains of mice that are slightly older than one another? It is possible that 
there could have been confounding factors, such as changes in enviromental 
conditions such as temperature or feeding regime, that may have effected gene
expression. In which case it was important to set the experiment up with 
control for both time points.
>
> 2. Extract the results for d33 v d11 for Infected mice. 
The the intercept is Uninfected mice at 11 days post infection, so the main 
effect `TimePoint_d33_vs_d11` is the result that we want.
```{r}
results.d33_v_d11_infected <- results(ddsObj.interaction, 
      contrast = list(c("TimePoint_d33_vs_d11", "TimePointd33.StatusInfected")),
                                        alpha = 0.05)
```
> How many genes have an adjusted p-value less than 0.05?  
```{r}
table(results.d33_v_d11_infected$padj < 0.05)
```
>
> Do these results suggest another approach to analysing this data set?   
Could we possibly treat the six uninfected samples as a single group with six
replicates and then just have 1 factor with 3 levels: Control, d11.Infected,
d33.Infected? This is really a biological question and not a statistical one.