---
title: "Introduction to Gene Set Testing in R"
date: "April 2021"
output:
  ioslides_presentation:
    css: css/stylesheet.css
    logo: images/CRUK_Cambridge_Institute.png
    smaller: yes
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
---


```{r setup, include=FALSE}
library(tidyverse)
library(gganimate)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  gganimate = list(
    nframes = 50
  ),
  out.width = '100%',
  dev.args=list(bg="transparent")
)
theme_set(theme_bw(base_size=12)) 
theme_update(panel.background = element_rect(fill = "transparent", colour = NA), 
             plot.background = element_rect(fill = "transparent", colour = NA)) 
```

## Differential Gene Expression Analysis Workflow {#less_space_after_title}

<div style="line-height: 50%;"><br></div>

<img src="images/workflow_3Day.svg" class="centerimg" style="width: 80%; margin-top: 60px;">
<!-- <img src="images/workflow.svg" class="centerimg" style="width: 60%;"> -->

## Gene Set Testing

There are many approaches to searching for biological meaning in the results
of differential expression analysis.

Commonly we look to see if the differentially expressed genes tend to relate
to specific pathways or ontological groups of genes.

We will look at two methods of doing this:

* Over Representation Analysis (ORA)

* Gene Set Enrichment Analysis (GSEA)

## Gene Set Testing

Common sources of gene sets:

* KEGG pathways

* Gene Ontologies

* Reactome

* MSigDB (GSEA)

* Manually curated gene lists

## Over Representation Analysis - Method

* This method tests whether genes in specific pathway are present in a subset of 
  genes of interest in our data more than expected.

* The genes of interest could be e.g. statistically significant 
  genes or a cluster of genes from hierachical or k-means clustering.
  
* Given the ratio of genes in the pathway to genes not in the pathway, is the
  number of genes in the pathway and in our subset statistically unlikely by 
  chance.

<img src="images/OverRepresentationAnalysis.svg" class="centerimg" style="width: 90%; margin-top=10px">

## Over Representation Analysis - Method

Genes in the experiment are split in two ways:
  
* annotated to the pathway or not
* differentially expressed or not

Contingency table:
  
<img src="images/ContingencyTable.svg" class="centerimg" style="width: 45%">

* Analysis with the hypergeometric/Fisher's exact test


## Gene Set Enrichment Analysis (GSEA)

* This method is based on ranking of all genes in our dataset

* If the gene set is significantly affected in our experiment, then the genes
  in the set should tend to be at one end or the other of our ranking.

* The ranking method is arbitrary, but p-value and fold change are common
  choices.
  
* GSEA calculates an enrichment score based on the ranking, and then uses 
  permutation to calculate a p-value for how significant the enrichment score   is.

## GSEA: Calculate the enrichment score

* Ranking by Fold Change 
```{r loadGSEAdata, include=FALSE, message=FALSE}
set.seed(1)
tab <- readRDS("RObjects/Shrunk_Results.d11.rds") %>% 
    slice(1:150) %>% 
    select(logFC) %>% 
    rowid_to_column("Index") %>% 
    arrange(desc(logFC)) %>% 
    rowid_to_column("Index2") %>% 
    mutate(inGeneSetA=Index2%in%c(2,3,6,8,
                                  10,13,19,21,
                                  11,28,12,35,
                                  24,41,90,100,
                                  110,125,136,148)) %>%  
    mutate(Score=ifelse(inGeneSetA, 8/20, -8/130)) %>% 
    mutate(CumScoreA=cumsum(Score)) %>% 
    mutate(inGeneSetB=Index2%in%sample(1:150, 20, replace = FALSE)) %>%  
    mutate(Score=ifelse(inGeneSetB, 8/20, -8/130)) %>% 
    mutate(CumScoreB=cumsum(Score))
```

<div style="width: 80%; margin-left: 13%">
```{r Foldchanges, echo = FALSE, dpi = 300, fig.width= 4, fig.height=2.5}
tab %>% 
    ggplot(aes(x=Index, y=logFC)) +
        geom_segment(aes(xend=Index), yend=0)
```
</div> 

## GSEA: Calculate the enrichment score

* Ranking by Fold Change 

<div style="width: 80%; margin-left: 13%">
```{r FoldchangesSorted, echo = FALSE, dpi = 300, fig.width= 4, fig.height=2.5}
tab %>% 
    ggplot(aes(x=Index2, y=logFC)) +
    geom_segment(aes(xend=Index2), yend=0) +
    labs(x="Ranking Index")
```
</div>

## GSEA: Calculate the enrichment score

* Identify genes in set

<div style="width: 80%; margin-left: 13%">
```{r FoldchangesWithGeneSet, echo = FALSE, dpi = 300, fig.width= 4, fig.height=2.5}
tab %>%  
    ggplot(aes(x=Index2, y=logFC)) +
    geom_segment(aes(xend=Index2, colour=inGeneSetA), yend=0) +
    scale_colour_manual(values=c("darkgrey", "red")) +
    guides(colour=FALSE) +
    labs(x="Ranking Index")
```
</div>

## GSEA: Calculate the enrichment score

* Calculate the enrichment score ... start at 0 and an enrichment score 
of 0

<div style="width: 80%; margin-left: 13%">
```{r Ranking, echo = FALSE, dpi = 300, fig.width= 4, fig.height=2.5}
tab %>%  
    ggplot(aes(x=Index2)) +
    geom_segment(aes(xend=Index2, colour=inGeneSetA), y=-0.5, yend=0.5) +
    geom_point(x=0, y=0, size=1, colour = "black") +
    scale_y_continuous(limits=c(-2,4)) +
    labs(x="Ranking Index", y="Enrichment Score") +
    scale_colour_manual(values=c("darkgrey", "red")) +
    guides(colour=FALSE)
```
</div>

## GSEA: Calculate the enrichment score

* Walk along genes and calculate a cumulative score

<div style="width: 80%; margin-left: 13%">
```{r anim1, echo=FALSE, message=FALSE, warning=FALSE}
p <- tab %>% 
    ggplot() +
    geom_segment(aes(x=Index2, xend=Index2, colour=inGeneSetA, 
                     group=seq_along(Index2)), 
                  y=-0.5, yend=0.5) +
    geom_line(aes(x=Index2, y=CumScoreA)) +
    scale_y_continuous(limits=c(-2,4)) +
    scale_colour_manual(values=c("darkgrey", "red")) +
    guides(colour=FALSE) +
    labs(x="Ranking Index", y="Enrichment Score") +
    transition_reveal(Index2)

animate(p, nframes=150, fps = 8, renderer = gifski_renderer(loop=FALSE), 
        res = 300, width= 4, height=2.5, units="in", bg="transparent")
```
</div>

## GSEA: Calculate the enrichment score

* A different gene set

<div style="width: 80%; margin-left: 13%">
```{r genesetB, echo=FALSE, dpi = 300, fig.width= 4, fig.height=2.5}
tab %>% 
    ggplot(aes(x=Index2)) +
    geom_segment(aes(xend=Index2, colour=inGeneSetB), y=-0.5, yend=0.5) +
    scale_y_continuous(limits=c(-2,4)) +
    labs(x="Ranking Index", y="Enrichment Score") +
    scale_colour_manual(values=c("darkgrey", "red")) +
    guides(colour=FALSE)
```
</div>


## GSEA: Calculate the enrichment score

* A different gene set

<div style="width: 80%; margin-left: 13%">
```{r anim2, echo=FALSE}
p <- tab %>% 
    ggplot(aes(x=Index2)) +
    geom_segment(aes(xend=Index2, colour=inGeneSetB,
                     group=seq_along(Index2)), 
                  y=-0.5, yend=0.5) +
    geom_line(aes(y=CumScoreB)) +
    scale_y_continuous(limits=c(-2,4)) +
    labs(x="Ranking Index", y="Enrichment Score") +
    scale_colour_manual(values=c("darkgrey", "red")) +
    guides(colour=FALSE) +
    transition_reveal(Index2)

animate(p, nframes=75, fps = 8, renderer = gifski_renderer(loop=FALSE), 
        res = 300, width= 4, height=2.5, units="in", bg="transparent")
```
</div>

## GSEA - estimate a p-value

* Randomly permute the ranking and recalculate the enrichment score.

* From a distribution of our permuted enrichment scores determine how likely our ES.

<div style="width: 60%; margin-left: 20%; padding-top: 30px">
```{r pvalue, echo=FALSE}
x1 <- seq(-6, 6, length=100)
hx1 <- dnorm(x1, mean = 0, sd = 1.5)
par(bg=NA, mar=c(5, 4, 0, 4) + 0.1) 

plot(x1, hx1, type="l", lty=1, lwd=2,
     xlab="Enrichment Score", ylab="Density",
     col="tomato", ylim=c(0, 0.3), xlim=c(-7, 7))
abline(v=4, col="darkgreen", lty=2, lwd=2)
```
</div>
