---
title: "Introduction to Bulk RNAseq data analysis"
subtitle: "A very brief intro to `dplyr` and tidyverse"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
minutes: 300
layout: page
bibliography: ref.bib
editor_options: 
  chunk_output_type: inline
---

One of the most complex aspects of learning to work with data in `R` is 
getting to grips with subsetting and manipulating data tables. The package 
`dplyr` [@Wickham2018] was developed to make this process more intuitive than it
is using standard base `R` processes. It also makes use of a new symbol `%>%`,
called the "pipe", which makes the code a bit tidier. 

`dplyr` is one of suite of similar packages collectively known as the tidyverse.

This is a **very** brief introduction to the `tidyverse` way of writing R code.
A more detailed introduction can be found in our
[online R course](https://bioinformatics-core-shared-training.github.io/r-intro/week4.html) 

We are introducing this because it makes many of the processes we will look at 
later **much simpler**. Importantly it also results in code that is much easier
to read and understand.

The entire tidyverse suite can be loaded via the `tidyverse` package:

```{r setup}
library(tidyverse)
```

Let's have a quick look at this by playing with our `sampleinfo` table.

```{r loadSampleInfo, message = FALSE}
# Read the sample information into a data frame
sampleinfo <- read_tsv("data/samplesheet.tsv")
sampleinfo
```

Suppose we wanted a new sample table that:

1. Just includes the "d11" samples
2. Only has the columns "SampleName" and "Status"
3. Renames the "Status" column as "SampleGroup"

### Manipulating the table in base R

With base R we would do something like this

```{r baseR, }
newTable <- sampleinfo
newTable <- newTable[newTable$TimePoint=="d11",]
newTable <- newTable[, c("SampleName", "Status")]
colnames(newTable)[2] <- "SampleGroup"
newTable
```

### `dplyr`

With `dplyr` we can use three new functions: `filter`, `select` and `rename`:

```{r dplyr}
newTable <- sampleinfo
newTable <- filter(newTable, TimePoint=="d11")
newTable <- select(newTable, SampleName, Status)
newTable <- rename(newTable, SampleGroup=Status)
newTable
```

The idea is that the `dplyr` code is easier to read and interpret than the base 
R syntax. 

There's no need to quote the column names as dplyr intelligently interprets
the arguments it's passed as belonging to the data table columns.

### The Pipe

Rather than repeatedly reassigning `newTable <- f(newTable)` as above, we can 
use the "*pipe*" - `%>%`. This takes the output of one function and "*pipes*" it
into the first argument of the next function so that we don't have to keep
specifying the object we are working with:

```{r pipe}
newTable <- sampleinfo %>%
    filter(TimePoint=="d11") %>%
    select(SampleName, Status) %>% 
    rename(SampleGroup=Status)
newTable
```

This is a fairly trivial example and the benefits may not be immediately 
obvious, but once you get used to using `dplyr` (and the other related
"tidyverse" packages, such as `stringr`) you'll find it much more powerful and 
easy to use than base R. 

----------------------------------------------

# References