---
title: "Introduction to Bulk RNAseq data analysis"
subtitle: Annotation Differential Expression Results with biomaRt
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
layout: page
bibliography: ref.bib
---

```{r setup, message=FALSE}
library(biomaRt)
library(tidyverse)
```

```{r knitrOpts, echo=FALSE}
knitr::opts_knit$set(cache=TRUE)
```

# Overview

- Getting annotation

# Adding annotation to the DESeq2 results

We have a list of significantly differentially expressed genes, but the only
annotation we can see is the Ensembl Gene ID, which is not very informative. 

```{r, message=FALSE}
results.interaction.11 <- readRDS("RObjects/DESeqResults.interaction_d11.rds")
results.interaction.11
```


There are a number of ways to add annotation. One method is to do this using the
*org.Mm.eg.db* package. This package is one of several *organism-level* packages
which are re-built every 6 months. These packages are listed on the [annotation 
section](http://bioconductor.org/packages/release/BiocViews.html#___AnnotationData) 
of the Bioconductor, and are installed in the same way as regular Bioconductor 
packages. 

An alternative approach is to use `biomaRt`, an interface to the 
[BioMart](http://www.biomart.org/) resource. This is the method we will use 
today.

## Select BioMart database and dataset

The first step is to select the Biomart database we are going to access and 
which data set we are going to use.

There are multiple mirror sites that we could use for access. The default is to 
use the European servers, however if the server is busy or inaccessible for some
reason it is possible to access one of the three mirror sites. See the 
instructions at 
[here](https://bioconductor.org/packages/release/bioc/vignettes/biomaRt/inst/doc/accessing_ensembl.html) 
for detailed instruction on using different mirrors, but in brief simply add the 
`mirror` argument to the `listEnsembl` and `useEnsembl` functions below.   

e.g to use the US West mirror:  
  `ensembl <- useEnsembl("genes", mirror = "uswest")`  
  
**Pro Tip**: The Ensembl servers tend to get very busy in the afternoons, to 
the extent that biomaRt may have trouble getting and maitaining a connection. 
Try to do this in the morning.

### list the available databases

```{r ensDBs}
# view the available databases
listEnsembl()
```
```{r echo=FALSE, include=FALSE}
# view the available databases
currentRelease <- listMarts()[1,2] %>% str_extract("[0-9]+$")
```

We want the "genes" database.

The current Ensembl release is version **`r currentRelease`**, however, when we 
downloaded the reference files the current version was **102**. We will need to
ensure that `biomaRt` accesses the correct version of the gene annotations.

```{r ensArchivedDBs}
# view the available databases
listEnsembl(version = 102)
```


### list the available datasets (species)

Now we need to identify the dataset for mouse.

```{r datasets}
## set up connection to ensembl database
ensembl <- useEnsembl(biomart = "genes", version=102)

# serach the available datasets (species)
searchDatasets(mart = ensembl, pattern = "Mouse")
```

We want the `mmusculus_gene_ensembl` data set.

### specify a data set to use

We can now specify the complete connection to the database and dataset that we
want to access.

```{r connect}
ensembl <- useEnsembl(biomart = 'genes', 
                      dataset = 'mmusculus_gene_ensembl',
                      version = 102)
```

## Query the database

Now we need to set up a query. For this we need to specify three things: 

(a) What type of information we are going to search the dataset on - called
**filters**. In our case this is Ensembl Gene IDs
(b) A vector of the **values** for our filter - the Ensembl Gene IDs from our DE 
results table
(c) What columns (**attributes**) of the dataset we want returned.


```{r filters, message=F}
# check the available "filters" - things you can filter for
ensembl_filters  <- listFilters(ensembl)

# To find the correct name for the Ensembl ID we can filter the name column
ensembl_filters %>% 
    filter(str_detect(name, "ensembl"))
```

So, we will use `ensembl_gene_id` to query the data bases

```{r attributes, message=F}
# check the available "attributes" - things you can retreive
ensembl_attributes <- listAttributes(ensembl)
head(ensembl_attributes, 20)
```

We'll retrieve the `external_gene_name`, which is the Gene Symbol, the 
`entrez_id` - we may need this for tools that use the NCBI databases - 
and the `entrez_accession`, which is the Gene Symbol associated with that 
`entrez_id`.

We also need to specify that we want the query to return the `ensembl_gene_id`
that we used to query the database. This ensures that our final table will 
include all of query gene ids, even if no record was found in the data base.

Returning data from Biomart can take time, so it's always a good idea to test 
your query on a small list of values first to make sure it is doing what you
want. We'll just use the first 1000 genes for now.

```{r runQuery}
# Set the filter type and values
ourFilterType <- "ensembl_gene_id"

# get the Ensembl IDs from our results table
filterValues <- rownames(results.interaction.11)[1:1000]

# Set the list of attributes
attributeNames <- c("ensembl_gene_id",
                    "external_gene_name", 
                    "entrezgene_id", 
                    "entrezgene_accession")

# run the query
annot <- getBM(attributes=attributeNames, 
               filters = ourFilterType, 
               values = filterValues, 
               mart = ensembl)
head(annot)
```

### One-to-many relationships

Let's inspect the annotation.

```{r inspectAnnot}
dim(annot)
```

#### Why are there more than `r nrow(annot)` annotations?

Some genes that have multiple entries in the retrieved annotation. This is 
because there are multiple Entrez IDs for a single Ensembl gene. These 
one-to-many relationships come up frequently in genomic databases, it is 
important to be aware of them and check when necessary. 

```{r getDups}
# find all rows containing duplicated ensembl ids
annot %>%  
    add_count(ensembl_gene_id) %>%  
    filter(n>1)
```

We will need to do a little work before adding the annotation to our results 
table. We could decide to discard one or both of the Entrez ID mappings, or we 
could concatenate the Entrez IDs so that we don't lose information. Matching
the `entrezgene_accession` and the `external_gene_name` can help with resolving
some of these problems. 


```{r fixDups}
fixedDuplicates <- annot %>%  
    add_count(ensembl_gene_id) %>%  
    filter(n>1) %>% 
    select(-n) %>% 
    filter(entrezgene_accession==external_gene_name)

annot <- annot %>%  
    add_count(ensembl_gene_id) %>%  
    filter(n==1) %>% 
    select(-n) %>% 
    bind_rows(fixedDuplicates)

nrow(annot)
length(unique(annot$ensembl_gene_id))
```


## Retrieve full annotation

> ### Challenge 1 {.challenge}
> That was just 1000 genes. We need annotations for the entire results table.
> Also, there may be some other interesting columns in BioMart that we wish to
> retrieve.  
>
> (a) Search the attributes and add the following to our list of attributes:  
>       (i) The gene description   
>       (ii) The gene biotype  
> (b) Query BioMart using all of the genes in our results table (`results.interaction.11`)  
>
> (c) How many Ensembl genes have multipe Entrez IDs associated with them?  
> (d) How many Ensembl genes in `results.interaction.11` don't have any annotation?

-------------------------------------------------------------------------------